Single Site, Single Disturbance, MultiPool

### 2020-11-18 SingleSite_SingleDisturb.Rmd
Single Site Multi-Disturbance Multi-Pool

plot(time,n,type='l',ylim=range(n),xlim=xlim,xlim=xlim,ylab="LAI",lwd=3)
obs.sd = 0.75
n.true=n
n.obs = rnorm(NT,n.true,obs.sd)
points(time,n.obs,pch=18,col=2,cex=1.5)
legend("topleft",legend=c("truth","obs"),lwd=c(2,NA),pch=c(NA,18),col=1:2)
## Simulate less frequent observations
#na.seq = 1:NT; na.seq=na.seq[-seq(1,NT,by=5)]
#n.fia = n.obs;n.fia[na.seq] <- NA
#points(time,n.fia,pch=18,col="blue",cex=2)
```

@@

```{r}
n0.sd = n.sd
NE = 1000
n0s = rlnorm(NE,log(nIC),n0.sd)
ne0 = matrix(n0s,NE,NT)
for(i in 1:NE){
  for(t in 2:NT){
    ne0[i,t] = abs(ne0[i,t-1] + r*ne0[i,t-1]*(1-ne0[i,t-1]/K) + rnorm(1,0,n.sd))
#    if(rbinom(1,1,p.dist)) ne[i,t] = abs(rnorm(1,n0,0.5))
  }
}
yrng = c(0,8) #range(ne0)
n0.stats = apply(ne0,2,quantile,c(0.025,0.5,0.975))
plot(time,n.obs,type='n',ylim=yrng,xlim=xlim)
polygon(cbind(c(time, rev(time), 1), c(n0.stats[1, ], rev(n0.stats[3, ]),
  n.stats[1, 1])), border = NA, col = 'lightblue')
lines(time,n0.stats[2,],lwd=2,col="blue4")
points(time,n.obs,pch=18,col=2,cex=1.5)
```

@@

### Comparision to EnKF
```{r}
plot(data$Y,ylim = range(rbind(CIN,data$Y)),xlim=xlim,type='n',
     ylab="LAI",xlab="time")
polygon(cbind(c(time, rev(time), 1), c(an0.stats[2, ], rev(an0.stats[4, ]),
  an0.stats[1, 1])), border = NA, col = '#FFBF6C')
polygon(cbind(c(time, rev(time), 1), c(an.stats[2, ], rev(an.stats[4, ]),
  an.stats[1, 1])), border = NA, col = ecoforecastR::col.alpha('#b2df8a',0.7))
#ecoforecastR::ciEnvelope(1:NT,CIN[1,],CIN[3,],col=rgb(col[1],col[2],col[3],0.3*256,maxColorValue=256))
ecoforecastR::ciEnvelope(1:NT,CI.fN[2,],CI.fN[4,],col=rgb(col[1],col[2],col[3],0.3*256,maxColorValue=256))

lines(an0.stats[3,],col="orange",lwd=2)
lines(an.stats[3,],col="#b2df8a",lwd=2)
lines(CI.fSSSD[2,],col="blue",lwd=2)
lines(Dbar,lwd=2)
points(data$Y,pch=18,col=2)
legend("topright",legend=c("EnKF, no disturb","EnKF, disturb","Multinomial","data","P(dist)"),
       lwd=5,col=c("orange","#b2df8a","blue","red","black"))
```

>>>>

plot(time,n,type='l',ylim=range(n),xlim=xlim,ylab="LAI",lwd=3)
obs.sd = 0.75
n.true=n
n.obs = rnorm(NT,n.true,obs.sd)
points(time,n.obs,pch=18,col=2,cex=1.5)
legend("topleft",legend=c("truth","obs"),lwd=c(2,NA),pch=c(NA,18),col=1:2)
## Simulate less frequent observations
#na.seq = 1:NT; na.seq=na.seq[-seq(1,NT,by=5)]
#n.fia = n.obs;n.fia[na.seq] <- NA
#points(time,n.fia,pch=18,col="blue",cex=2)
```

@@

```{r}
n0.sd = n.sd
NE = 1000
n0s = rlnorm(NE,log(nIC),n0.sd)
ne0 = matrix(n0s,NE,NT)
for(i in 1:NE){
  for(t in 2:NT){
    ne0[i,t] = abs(ne0[i,t-1] + r*ne0[i,t-1]*(1-ne0[i,t-1]/K) + rnorm(1,0,n.sd))
#    if(rbinom(1,1,p.dist)) ne[i,t] = abs(rnorm(1,n0,0.5))
  }
}
yrng = c(0,8) #range(ne0)
n0.stats = apply(ne0,2,quantile,c(0.025,0.5,0.975))
plot(time,n.obs,type='n',ylim=yrng,xlim=xlim)
polygon(cbind(c(time, rev(time), 1), c(n0.stats[1, ], rev(n0.stats[3, ]),
  n0.stats[1, 1])), border = NA, col = 'lightblue')
lines(time,n0.stats[2,],lwd=2,col="blue4")
points(time,n.obs,pch=18,col=2,cex=1.5)
```

@@

### Comparision to EnKF
```{r}
plot(data$Y,ylim = range(rbind(CI.fN,data$Y)),xlim=xlim,type='n',
     ylab="LAI",xlab="time")
polygon(cbind(c(time, rev(time), 1), c(an0.stats[2, ], rev(an0.stats[4, ]),
  an0.stats[1, 1])), border = NA, col = '#FFBF6C')
polygon(cbind(c(time, rev(time), 1), c(an.stats[2, ], rev(an.stats[4, ]),
  an.stats[1, 1])), border = NA, col = ecoforecastR::col.alpha('#b2df8a',0.7))
#ecoforecastR::ciEnvelope(1:NT,CIN[1,],CIN[3,],col=rgb(col[1],col[2],col[3],0.3*256,maxColorValue=256))
ecoforecastR::ciEnvelope(1:NT,CI.fN[2,],CI.fN[4,],col=rgb(col[1],col[2],col[3],0.3*256,maxColorValue=256))

lines(an0.stats[3,],col="orange",lwd=2)
lines(an.stats[3,],col="#b2df8a",lwd=2)
#lines(CI.fSSSD[2,],col="blue",lwd=2)
lines(CI.fN[3,],col="blue",lwd=2)
lines(Dbar,lwd=2)
points(data$Y,pch=18,col=2)
legend("topright",legend=c("EnKF, no disturb","EnKF, disturb","Multinomial","data","P(dist)"),
       lwd=5,col=c("orange","#b2df8a","blue","red","black"))
```

### 2020-11-18 SingleSite_SingleDisturb_MultiPool.Rmd
Single Site Multi-Disturbance Multi-Pool

```{r}
## spin up
ne = 100
met = sample.int(nd,ne,replace=TRUE)
par = sample.int(np,ne,replace=TRUE)
Na <- matrix(NA,ne,3)
for(i in 1:ne){
  Na[i,] = VSEM(as.numeric(Theta[par[i],]),PAR = PAR[met[i],])[nt,2:4]
}
Na.orig = Na
Nmu = NmuK = NfK = Nf = Na*NA
par(mfrow=c(1,3))
hist(Na[,1],main="Cleaf")
hist(Na[,2],main="Csoil")
hist(Na[,3],main="Cstem")
```

@@

  ## Forecast 
  for(i in 1:ne){
    Nmu[i,]  = VSEM(THETA[i,], PAR = PAR[met[i],tsel])[365,2:4] ## deterministic simulation
    NmuK[i,] = VSEM(THETAK[i,],PAR = PAR[met[i],tsel])[365,2:4] ## deterministic simulation
    epsilon  = mvtnorm::rmvnorm(1,rep(0,3),Q)                   ## process error  ** shared between versions **
    Nf[i,]   = Nmu[i,]  + epsilon
    NfK[i,]  = NmuK[i,] + epsilon
  }

@@

  ## update parameters & save Analysis posterior
  dat = as.matrix(jdat)
  nsel <- grep(pattern="^n",x=colnames(dat))
  Analysis[[t]] <- dat

  ## draw new IC
  Na <- multi.ens.adj(
          Xf   = Nf,
          cf   = d.prior,
          mu.f = rbind(priors$mufN,priors$mufD),
          Pf   = abind::abind(solve(priors$pfN),
                 solve(priors$pfD),
                 along=3),
          Xa = dat[,nsel],
          ca = dat[,"D"])
  Ca <- Na[,1] ## updated class assignments
  Na <- tobit(Na[,-1])
#  sel.ic <- sample.int(nrow(dat), ne, replace=TRUE)
#  Na <- dat[sel.ic,nsel]

@@

```{r}
pool.names=c("Cleaf","Csoil","Croot")
for(i in 1:3){

  CI.Na = sapply(X = Analysis,function(x){
    sel.n = which(colnames(x) == paste0("n[",i,"]"))
    quantile(x[,sel.n],c(0.025,0.5,0.975))
  })
  #CI.Na = sapply(X = Analysis,function(x){
  #  apply(x,2,quantile,c(0.025,0.975))
  #})

  plot(No[,i],
       #ylim = range(rbind(CIN,data$Y)),xlim=xlim,type='n',
       ylab=pool.names[i],xlab="time")
  ecoforecastR::ciEnvelope(1:ncol(CI.Na),CI.Na[1,],CI.Na[3,],col=ecoforecastR::col.alpha('#b2df8a',0.7))
  lines(CI.Na[2,],col="#b2df8a",lwd=2)
  lines(Ntrue[,i],lty=2,lwd=2)
}
```

## EnKF, no disturbance

>>>>

```{r}
## spin up
ne = 200
met = sample.int(nd,ne,replace=TRUE)
par = sample.int(np,ne,replace=TRUE)
Na <- matrix(NA,ne,3)
for(i in 1:ne){
  Na[i,] = VSEM(as.numeric(Theta[par[i],]),PAR = PAR[met[i],])[nt,2:4]
}
Na.orig = Na
Nmu = NmuK = NfK = Nf = Na*NA
par(mfrow=c(1,3))
hist(Na[,1],main="Cleaf")
hist(Na[,2],main="Csoil")
hist(Na[,3],main="Cstem")
```

@@

  ## Forecast 
  for(i in 1:ne){
    fx       = VSEM(THETA[i,], PAR = PAR[met[i],tsel]) ## deterministic simulation
    Nmu[i,]  =  fx[365,2:4]                         ## state variables at time of assim
    fxK      = VSEM(THETAK[i,],PAR = PAR[met[i],tsel]) ## deterministic simulation
    NmuK[i,] = fxK[365,2:4] ## deterministic simulation
    epsilon  = mvtnorm::rmvnorm(1,rep(0,3),Q)                   ## process error  ** shared between versions **
    Nf[i,]   = Nmu[i,]  + epsilon
    NfK[i,]  = NmuK[i,] + epsilon
  }

@@

  ## update parameters & save Analysis posterior
  dat = as.matrix(jdat)
  nsel <- grep(pattern="^n",x=colnames(dat))

  ## draw new IC
  Na <- multi.ens.adj(
          Xf   = Nf,
          cf   = d.prior,
          mu.f = rbind(priors$mufN,priors$mufD),
          Pf   = abind::abind(solve(priors$pfN),
                 solve(priors$pfD),
                 along=3),
          Xa = dat[,nsel],
          ca = dat[,"D"])
  Ca <- Na[,1] ## updated class assignments
  Na <- tobit(Na[,-1])
#  sel.ic <- sample.int(nrow(dat), ne, replace=TRUE)
#  Na <- dat[sel.ic,nsel]
  Analysis[[t]] <- list(MCMC=dat,
                        Na = Na,
                        Ca = Ca)

@@

### time series

```{r}
pool.names=c("Cleaf","Csoil","Cstem")
for(i in 1:3){

  CI.Na = sapply(X = Analysis,function(y){
    x = y$MCMC
    sel.n = which(colnames(x) == paste0("n[",i,"]"))
    quantile(x[,sel.n],c(0.025,0.5,0.975))
  })
  #CI.Na = sapply(X = Analysis,function(x){
  #  apply(x,2,quantile,c(0.025,0.975))
  #})

  plot(No[,i],
       #ylim = range(rbind(CIN,data$Y)),xlim=xlim,type='n',
       ylab=pool.names[i],xlab="time")
  ecoforecastR::ciEnvelope(1:ncol(CI.Na),CI.Na[1,],CI.Na[3,],col=ecoforecastR::col.alpha('#b2df8a',0.7))
  lines(CI.Na[2,],col="#b2df8a",lwd=2)
  lines(Ntrue[,i],lty=2,lwd=2)
}
```
### Ensemble adj at time of disturbance

```{r,fig.asp=1,fig.width=3}
#par(mfrow=c(1,3))

plot(Analysis[[10]]$Na[,1],Analysis[[10]]$Na[,3],pch="+",col=ecoforecastR::col.alpha("red",alpha=0.5),
     ylim=range(Forecast[[10]]$Nf[,3]),
     xlim=range(Forecast[[10]]$Nf[,1]),
     xlab=pool.names[1],ylab=pool.names[3],
     cex.lab=1.25,asp = 1)
points(Forecast[[10]]$Nf[,1],Forecast[[10]]$Nf[,3],cex=0.5)
points(No[10,1],No[10,3],cex=3,col=ecoforecastR::col.alpha("purple",alpha = 0.75),pch=18)

## ensemble adjustment
sel.a=sample.int(ne,10)
arrows(Forecast[[10]]$Nf[sel.a,1],Forecast[[10]]$Nf[sel.a,3],
       Analysis[[10]]$Na[sel.a,1],Analysis[[10]]$Na[sel.a,3],length = 0.1,lwd = 0.5)   
legend("topleft",
       legend=c("Forecast","Analysis","Data"),
       pch=c(1,3,18),col=c(1,2,"purple"),
       cex = 1.5)

```
## EnKF, no disturbance

### 2020-11-18 methods.Rmd
Single Site Multi-Disturbance Multi-Pool

@@

3. Simulated data experiments (known disturbance)
  + Single Site, Single Disturbance
  + SSSD Multi Pool
  + SS Multi-Disturbance w/ PFT switching
  + Simple Multi-site: spatial cov
4. Real-world examples
  + Describe Model & PEcAn
  + Calibrate state-and-transition prior
    + Landtrendr
    ftp://islay.ceoas.oregonstate.edu/change_agent_attribution
  + Applications: PNW
    + Logging
    + Fire
    + Land use transition (forest -> non?)

>>>>

*Priors*

  $$X^L_k \sim MVN(\vec{\mu}_{f,k},P_{f,k})$$
  $$\vec{\rho} \sim Multinomial(\vec{p})$$ 
*Process Model*

  $$X = X^L \vec{\rho}$$

*Likelihood*
  $$Y \sim MVN(X,R)$$
### Ensemble Adjustment

[@Anderson2001]

@@

3. Simulated data experiments (known disturbance)
  + x Single Site, Single Disturbance  (add a case where EnKF has disturb but it's more rare)
  + x SSSD Multi Pool
  + x SS Multi-Disturbance - distinct
  + SS Multi-Disturbance w/ distinct, PFT switching
  + SSMD - initially overlapping but diverge
      * how to handle updating of assignment based on new information?
  + SSMD - transition probabilities change with state
  + Simple Multi-site: spatial cov
  + rate of false pos/neg: affected by p.dist, R, Q, P0 -- how much do we need to simulate?
4. Real-world examples
  + Describe Model & PEcAn
  + Calibrate state-and-transition prior
    + Landtrendr
    ftp://islay.ceoas.oregonstate.edu/change_agent_attribution
  + Applications: PNW
    + Logging
    + Fire
    + Land use transition (forest -> non?)
